
title: Analysis of the human embryonic stem cell (hESC) dataset generated with Smart-seq2 using Bioconductor package simple SingleCellExperiment
Author: Saifun Nahar, Mohammad Suleman, Arman Moshasha
bibliography: references.bib
format:
  gfm:
    output-file: README.md
    execute:
      echo: false
      eval: true
  revealjs:
    slide-level: 4
    execute:
      echo: false
      eval: true

# Introduction
Smart-seq2 is the sequencing method used for the Messmer human embryonic stem cell (hESC) dataset. The analysis is performed using the Bioconductor package simple SingleCellExperiment, which is designed for handling single-cell RNA sequencing data. The workflow includes several steps such as data preprocessing, quality control, normalization, dimensionality reduction, clustering, and differential expression analysis. The goal of the analysis is to identify distinct cell populations and understand their gene expression profiles in the context of human embryonic stem cells.

# Data loading
The dataset is loaded into R using the `readRDS` function, which reads the serialized R object containing the single-cell data. The data is stored in a SingleCellExperiment object, which allows for efficient handling and analysis of single-cell RNA-seq data.

```{r}
update.packages(ask = FALSE)
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install(version = "3.16") 
BiocManager::install("scRNAseq")
```
```{r}
library(scRNAseq)
sce.mess <- MessmerESCData()
sce.mess$`experiment batch` <- factor(sce.mess$`experiment batch`)
```

```{r}
library(AnnotationHub)
ens.hs.v97 <- AnnotationHub()[["AH73881"]]
anno <- select(ens.hs.v97, keys=rownames(sce.mess), 
    keytype="GENEID", columns=c("SYMBOL"))
rowData(sce.mess) <- anno[match(rownames(sce.mess), anno$GENEID),]
```

# Quality control
Quality control is performed to filter out low-quality cells and genes. This typically involves calculating metrics such as the number of detected genes, total counts per cell, and the percentage of mitochondrial gene expression. Cells that do not meet certain thresholds for these metrics are removed from the analysis to ensure that only high-quality data is used for downstream analysis.

```{r}
colSums(as.matrix(filtered))
```

```{r}
gridExtra::grid.arrange(
    plotColData(original, x="experiment batch", y="sum",
        colour_by=I(filtered$discard), other_field="phenotype") +
        facet_wrap(~phenotype) + scale_y_log10(),
    plotColData(original, x="experiment batch", y="detected",
        colour_by=I(filtered$discard), other_field="phenotype") +
        facet_wrap(~phenotype) + scale_y_log10(),
    plotColData(original, x="experiment batch", y="subsets_Mito_percent",
        colour_by=I(filtered$discard), other_field="phenotype") +
        facet_wrap(~phenotype),
    plotColData(original, x="experiment batch", y="altexps_ERCC_percent",
        colour_by=I(filtered$discard), other_field="phenotype") +
        facet_wrap(~phenotype),
    ncol=1
)
```
# Normalization
Normalization is performed to account for differences in sequencing depth and other technical variations across cells. This step typically involves scaling the counts to a common library size and applying a transformation to stabilize the variance. Normalization allows for more accurate comparisons of gene expression levels across cells and is essential for downstream analyses such as clustering and differential expression.

```{r}
library(scran)

set.seed(10000)
clusters <- quickCluster(sce.mess)
sce.mess <- computeSumFactors(sce.mess, cluster=clusters)
sce.mess <- logNormCounts(sce.mess)

par(mfrow=c(1,2))
plot(sce.mess$sum, sizeFactors(sce.mess), log = "xy", pch=16,
     xlab = "Library size (millions)", ylab = "Size factor",
     col = ifelse(sce.mess$phenotype == "naive", "black", "grey"))

spike.sf <- librarySizeFactors(altExp(sce.mess, "ERCC"))
plot(sizeFactors(sce.mess), spike.sf, log = "xy", pch=16,
     ylab = "Spike-in size factor", xlab = "Deconvolution size factor",
     col = ifelse(sce.mess$phenotype == "naive", "black", "grey"))
```
# Cell cycle phase assignment
Cell cycle phase assignment is performed to classify cells into different phases of the cell cycle (G1, S, G2/M) based on the expression of known cell cycle marker genes. This can provide insights into the proliferative state of the cells and help to identify potential sources of variation in gene expression that are related to cell cycle effects.

```{r}
set.seed(10001)
hs_pairs <- readRDS(system.file("exdata", "human_cycle_markers.rds", package="scran"))
assigned <- cyclone(sce.mess, pairs=hs_pairs, 
    gene.names=rownames(sce.mess),
    BPPARAM=BiocParallel::MulticoreParam(10))
sce.mess$phase <- assigned$phases

table(sce.mess$phase)

smoothScatter(assigned$scores$G1, assigned$scores$G2M, xlab="G1 score",
     ylab="G2/M score", pch=16)
```

# Feature selection
Feature selection is performed to identify highly variable genes that are informative for downstream analyses such as clustering and dimensionality reduction. This typically involves calculating the variance of gene expression across cells and selecting genes that exhibit higher variability than expected based on their mean expression levels. The selected genes are then used for subsequent analyses to capture the biological heterogeneity in the dataset.

```{r}
dec <- modelGeneVarWithSpikes(sce.mess, "ERCC", block = sce.mess$`experiment batch`)
top.hvgs <- getTopHVGs(dec, prop = 0.1)
par(mfrow=c(1,3))
for (i in seq_along(dec$per.block)) {
    current <- dec$per.block[[i]]
    plot(current$mean, current$total, xlab="Mean log-expression", 
        ylab="Variance", pch=16, cex=0.5, main=paste("Batch", i))

    fit <- metadata(current)
    points(fit$mean, fit$var, col="red", pch=16)
    curve(fit$trend(x), col='dodgerblue', add=TRUE, lwd=2)
}
```
# Batch correction
Batch correction is performed to remove technical variations that arise from differences in experimental conditions or processing batches. This step is crucial for integrating data from multiple batches and ensuring that the observed biological differences are not confounded by batch effects. Methods such as ComBat, MNN (Mutual Nearest Neighbors), or Harmony can be used for batch correction, which adjust the gene expression values to mitigate the influence of batch-specific biases.

```{r}
library(batchelor)
sce.mess <- correctExperiments(sce.mess, 
    PARAM = RegressParam(
        design = model.matrix(~sce.mess$phenotype + sce.mess$`experiment batch`),
        keep = 1:2 
    )
)
```
# Dimensionality reduction
Dimensionality reduction techniques such as Principal Component Analysis (PCA) and t-distributed Stochastic Neighbor Embedding (t-SNE) are applied to the normalized data to visualize the high-dimensional gene expression data in a lower-dimensional space. This helps to identify patterns and clusters of cells based on their gene expression profiles. PCA reduces the dimensionality by finding the directions of maximum variance in the data, while t-SNE focuses on preserving local relationships between cells.

```{r}
set.seed(1101001)
sce.mess <- runPCA(sce.mess, subset_row = top.hvgs, exprs_values = "corrected")
sce.mess <- runTSNE(sce.mess, dimred = "PCA", perplexity = 40)
```
# From a naive PCA, the cell cycle appears to be a major source of biological variation within each phenotype. To mitigate this, we can regress out the cell cycle scores from the PCA.

```{r}
gridExtra::grid.arrange(
    plotTSNE(sce.mess, colour_by = "phenotype") + ggtitle("By phenotype"),
    plotTSNE(sce.mess, colour_by = "experiment batch") + ggtitle("By batch "),
    plotTSNE(sce.mess, colour_by = "CDK1", swap_rownames="SYMBOL") + ggtitle("By CDK1"),
    plotTSNE(sce.mess, colour_by = "phase") + ggtitle("By phase"),
    ncol = 2
)
```
# We perform contrastive PCA (cPCA) and sparse cPCA (scPCA) on the corrected log-expression data to obtain the same number of PCs. Given that the naive hESCs are actually reprogrammed primed hESCs, we will use the single batch of primed-only hESCs as the “background” dataset to remove the cell cycle effect.

```{r}
library(scPCA)
is.bg <- sce.mess$`experiment batch`=="3"
target <- sce.mess[,!is.bg]
background <- sce.mess[,is.bg]

mat.target <- t(assay(target, "corrected")[top.hvgs,])
mat.background <- t(assay(background, "corrected")[top.hvgs,])

set.seed(1010101001)
con_out <- scPCA(
    target = mat.target,
    background = mat.background,
    penalties = 0, # no penalties = non-sparse cPCA.
    n_eigen = 50,
    contrasts = 100
)
reducedDim(target, "cPCA") <- con_out$x

set.seed(101010101)
sparse_con_out <- scPCA(
    target = mat.target,
    background = mat.background,
    penalties = 1e-4,
    n_eigen = 50,
    contrasts = 100,
    alg = "rand_var_proj" # for speed.
)
reducedDim(target, "scPCA") <- sparse_con_out$x
```
# We see greater intermingling between phases within both the naive and primed cells after cPCA and scPCA.

```{r}
set.seed(1101001)
target <- runTSNE(target, dimred = "cPCA", perplexity = 40, name="cPCA+TSNE")
target <- runTSNE(target, dimred = "scPCA", perplexity = 40, name="scPCA+TSNE")

gridExtra::grid.arrange(
    plotReducedDim(target, "cPCA+TSNE", colour_by = "phase") + ggtitle("After cPCA"),
    plotReducedDim(target, "scPCA+TSNE", colour_by = "phase") + ggtitle("After scPCA"),
    ncol=2
)
```
# We can quantify the change in the separation between phases within each phenotype using the silhouette coefficient.

```{r}
library(bluster)
naive <- target[,target$phenotype=="naive"]
primed <- target[,target$phenotype=="primed"]

N <- approxSilhouette(reducedDim(naive, "PCA"), naive$phase)
P <- approxSilhouette(reducedDim(primed, "PCA"), primed$phase)
c(naive=mean(N$width), primed=mean(P$width))

cN <- approxSilhouette(reducedDim(naive, "cPCA"), naive$phase)
cP <- approxSilhouette(reducedDim(primed, "cPCA"), primed$phase)
c(naive=mean(cN$width), primed=mean(cP$width))

scN <- approxSilhouette(reducedDim(naive, "scPCA"), naive$phase)
scP <- approxSilhouette(reducedDim(primed, "scPCA"), primed$phase)
c(naive=mean(scN$width), primed=mean(scP$width))
```

# Reference:
Messmer, T., F. von Meyenn, A. Savino, F. Santos, H. Mohammed, A. T. L. Lun, J. C. Marioni, and W. Reik. 2019. “Transcriptional heterogeneity in naive and primed human pluripotent stem cells at single-cell resolution.” Cell Rep 26 (4): 815–24.